Open-source landscape for 3D CSEM modelling

### 2020-07-07 manuscript/Manuscript.tex
2nd round DW working on it

% \begin{equation}
% \begin{split}
%     \nabla \times \vec{E} &= -i \omega\vec{B}  \\
%     \nabla \times \vec{H} &= \vec{J}
% \end{split}
% \label{eq:maxwell_freq_quasi}
% \end{equation}

@@

\imycom[RR]{I think we should introduce the term degrees of freedom (dof),
  equivalent to the size of the SLE, as general measure of our problem sizes.
  If it is somewhat x*Nedges, or related to Ncells or Nnodes, who cares? We
  just need this one number for this paper in my opinion, more details in the
  jupyter notebooks - fine.}

The four codes under consideration are, in alphabetical order, \custem
\citep{GEO.19.Rochlitz}, \emg3d \citep{JOSS.19.Werthmuller}, \petgem
\citep{CAG.18.CastilloReyes, GJI.19.CastilloReyes}, and \simpeg
\citep{CAG.15.Cockett, CAG.17.Heagy}. All four codes have their user-facing
routines written
in Python; all of them make heavy use of \texttt{NumPy}
\citep{CSE.11.VanDerWalt} and \texttt{SciPy} \citep{NM.20.Virtanen}. The four
of them are “modern” open-source projects, meaning that they come with both, an
open-source license and an online-hosted version-control system with tracking
possibilities (raising issues, filing pull requests). All developments comprise
an extensive online documentation with many examples and have continuous
integration to some degree. Newer package-management systems such as Conda,
Docker, or pip significantly supported us to make our codes easily available
and installable without any user-side compilations. Each one can be downloaded
and installed with a single command.

In the following, we provide a quick overview of the codes. It is, however,
beyond the scope of this article to go into every detail of the different
modellers, and we refer to their documentations for more details. An overview
comparison of the different codes is given in Table~\ref{tbl:codecomparison}.
All codes have in common that they solve the weak formulation of Maxwell's
equation in its differential form (DE) under the quasistatic or diffusive
approximation, hence neglecting displacement currents. The machines on which
the different codes were run are listed in Table~\ref{tbl:machines} together
with the responsible operator.
% LH: SimPEG also interfaces to the scipy solvers; but in practice those are
%     use much less than Pardiso / mumps
% => I leave them out for now, as you did not use them for this paper either,
%    right? Maybe to mention in the SimPEG-section.

@@

%
\tabl[btp]{machines}{List of computer and operating system (hardware and
software) on which the different codes were run, together with the operator.}{
  \centering
  \footnotesize
  \begin{tabularx}{\linewidth}{lXl}
  \toprule
  %
  Code & Computer and Operating System & Operator \\
  \midrule
  %
  \custem & \textbf{PowerEdge R940 server}; 144 Xeon Gold 6154 CPU @2.666 GHz;
            ~3 TB DDR4 RAM; Ubuntu 18.04
          & Raphael Rochlitz \\
  \emg3d  & \textbf{Laptop}; i7-6600U CPU@2.6 GHz x4; 16 GB of memory,
            Ubuntu 18.04 & Dieter Werthmüller \\
  \petgem & \textbf{Marenostrum4 (supercomputer)}. Intel Xeon Platinum from
            Skylake generation; 2 sockets Intel Xeon Platinum 8160 CPU with 24
            cores each @2.10GHz for a total of 48 cores per node; 386 Gb DDR4
            RAM per node; SuSE Linux Enterprise
          & Octavio Castillo-Reyes \\
  \simpeg & ???
          & Lindsey Heagy \\
  %
  \bottomrule
\end{tabularx}}%
%

@@

The customisable electromagnetic modelling Python toolbox custEM was developed
for simulating arbitrary complex 3D CSEM geometries with focus on semi-airborne
setups, but it supports also land-based, airborne, coastal  or marine
environments. Multiple electric or magnetic field or potential finite-element
approaches were implemented as total or secondary field formulations. The
finite-element kernel, including higher order basis functions and
parallelisation, relies on the FEniCS project \citep{B.SPR.12.Logg,
B.SPR.16.Langtangen}. The resulting systems of linear equations are solved with
MUMPS \citep{SIAM.01.Amestoy}, which is a very robust but memory consuming
choice. Primary field solutions are supplied by the COMET \citep{GEO.20.Skibbe}
package.

@@

The 3D CSEM modeller \emg3d is a multigrid \citep{CMMP.64.Fedorenko} solver for
3D electromagnetic diffusion following \cite{GP.06.Mulder}, with tri-axial
electrical anisotropy, isotropic electric permittivity, and isotropic magnetic
permeability. The matrix-free solver can be used as main solver or as
preconditioner for one of the Krylov subspace methods implemented in SciPy, and
the governing equations are discretized on a staggered grid by the
finite-integration technique \citep{AEU.77.Weiland}, which is a finite-volume
generalization of a Yee grid. The code is written completely in Python using
the NumPy and SciPy stacks, where the most time- and memory-consuming parts are
sped up through jitted (just-in-time compiled) functions using Numba
\citep{LLVM.15.Lam}. The code computes the electric field due to an electric
source in the frequency-domain, which is its main application (frequency-domain
CSEM). However, there are also routines to obtain the magnetic field due to an
electric source, to obtain the electric and magnetic fields due to a magnetic
source, and to obtain time-domain responses.

The multigrid method is characterized by almost linear scaling both in terms of
runtime (CPU) and memory (RAM) usage, and it is therefore a comparably
low-memory consumption solver. It is also minimal in terms of requirements,
only NumPy, SciPy, and Numba are required, with Python 3.7+. The current
development is focused on adding basic inversion capabilities, and a further
plan is to implement a hook in order to use \emg3d as a frequency-domain CSEM
solver within the grander \simpeg framework.

@@

\petgem is a parallel code for frequency-domain 3D CSEM data from marine and
land surveys. The high-order edge FE method (HEFEM) is used for the
discretization of the governing equations in its diffusive form. This technique
provides a suitable mechanism to obtain stable numerical solutions as well as a
good trade-off between number of dof\mycom[OCR]{I assume we have previously
mentioned the term dofs} and computational effort. The current implementation
supports up to six-order tetrahedral vector basis functions. Moreover, because
the HEFEM belongs to the FE family, the unstructured meshes can be used
efficiently on complex geometries (e.g., models with topography and
bathymetry).

\petgem permits to locate the source and receivers anywhere in the
computational domain\mycom[DW]{I think this applies to all four codes} (e.g.,
sediments, seafloor, sea, ground), allowing to analyze the physical environment
of the electric responses and how parameters impact them (e.g., frequency,
conductivity, dependence on mesh setup, basis order, solver type, among
others). Furthermore, \petgem implements a semi-adaptive mesh strategy ($hp$
mesh refinement) based on physical parameters and on polynomial order to
satisfy quality criteria chosen by the user. Nonetheless, only Horizontal
Electric Dipole (HED) has been implemented.

For the parallel forward modelling computations, a highly scalable MPI domain
decomposition allows reducing run-times and the solution of large-scale
modelling cases. This strategy is capable of exploiting the parallelism offered
by both modest multi-core computers and cutting-edge clusters (e.g.,
High-performance Computing architectures).

@@

\simpeg is a modular toolbox for simulations and gradient based inversions in
geophysics. Current functionality includes modelling and inversion
capabilities for gravity, magnetics, direct current resistivity, induced
polarization, electromagnetics (time, frequency, as well as natural sources,
e.g. magnetotellurics), and fluid flow (Richards equation). It is a community
driven project that aims to support researchers by providing a flexible,
extensible framework and to facilitate integration of data sets from a variety
of geophysical methods, including joint inversions \citep{GJI.20.Astic}, by
providing a common interface to each method.

Meshes and finite volume differential operators are implemented in the
\discretize package. It currently includes tensor, OcTree, cylindrical, and
logically rectangular meshes. Each mesh type inherits from a common structure
and uses the same naming conventions for methods and properties. This allows us
to decouple the implementation of a discretized partial differential equation
from the details of the mesh geometry and therefore we can write a single
implementation of discretized partial differential equation (or set of partial
differential equations) in \simpeg that will support all mesh types implemented
in \discretize.

The electromagnetics implementations use a staggered grid finite volume
approach where physical properties are discretized at cell centers, fields on
cell edges and fluxes on cell faces \citep{IEEE.66.Yee} (for an overview of
discretizing Maxwell's equations, we recommend \cite{B.SIAM.14.Haber}). There are
two different discretization strategies implemented for Maxwell's equations: (
a) the EB-formulation, which discretizes the electric field ($\vec{e}$) on
cell edges and the magnetic flux density ($\vec{b}$) on cell faces, and (b)
the HJ-formulation, which discretizes the magnetic field ($\vec{h}$) on cell
edges and the current density ($\vec{j}$) on cell faces. In both cases the
physical properties: electrical conductivity / resistivity and magnetic
permeability are discretized at cell centers. Having multiple implementations
allows for testing that compares results from each approach, as well as the
representation of both electrical and magnetic sources on cylindrically
symmetric meshes. \simpeg supports variable magnetic permeability and
full-tensor anisotropy for the physical properties.

% Support for domain decomposition approaches (see for example
% \cite{GEO.20.Fournier})

@@

We computed the responses for two different models to verify that all different
3D codes yield the same electromagnetic responses. The first simple \emph{Block
model} consists of a layered, anisotropic background with three embedded
blocks. The second Marlim R3D model is based on a very complex, realistic
marine CSEM setup.

>>>>

% \begin{equation}
% \begin{split}
%     \nabla \times \vec{E} &= -i \omega\vec{B}  \\
%     \nabla \times \vec{H} &= \vec{J}
% \end{split}
% \label{eq:maxwell_freq_quasi}
% \end{equation}
%
% If we include Maxwell's equations, we might cite here Haber, 2014, for its
% discretization.

@@

The four codes under consideration are, in alphabetical order, \custem \citep{GEO.19.Rochlitz}, \emg3d \citep{JOSS.19.Werthmuller}, \petgem \citep{CAG.18.CastilloReyes, GJI.19.CastilloReyes}, and \simpeg \citep{CAG.15.Cockett, CAG.17.Heagy}. All four codes have their user-facing routines written in Python, and all of them make heavy use of \texttt{NumPy} \citep{CSE.11.VanDerWalt} and \texttt{SciPy} \citep{NM.20.Virtanen}. The four of them are “modern” open-source projects, meaning that they come with both, an open-source license and an online-hosted version-control system with tracking possibilities (raising issues, filing pull requests). All developments comprise an extensive online documentation with many examples and have continuous integration to some degree. Newer package-management systems such as Conda, Docker, or pip significantly supported us to make our codes easily available and installable on any of the major operating system without any user-side compilations. Each one can be downloaded and installed with a single command.

In the following, we provide a quick overview of the codes. It is, however, beyond the scope of this article to go into every detail of the different modellers, and we refer to their documentations for more details. An overview comparison of the different codes is given in Table~\ref{tbl:codecomparison}. All codes have in common that they solve the weak formulation of Maxwell's equation in its differential form (DE) under the quasistatic or diffusive approximation, hence neglecting displacement currents, which is a common approximation for the low frequency range commonly applied in CSEM. The machines on which the different codes were run are listed in Table~\ref{tbl:machines} together with the responsible operator.
% LH: SimPEG also interfaces to the scipy solvers; but in practice those are
%     use much less than Pardiso / mumps
% => I leave them out for now, as you did not use them for this paper either,
%    right? Maybe to mention in the SimPEG-section.

@@

%
\tabl[btp]{machines}{List of computer and operating system (hardware and
software) on which the different codes were run, together with the operator.}{
  \centering
  \footnotesize
  \begin{tabularx}{\linewidth}{lXl}
  \toprule
  %
  Code & Computer and Operating System & Operator \\
  \midrule
  %
  \custem & \textbf{PowerEdge R940 server}; 144 Xeon Gold 6154 CPU @2.666 GHz;
            ~3 TB DDR4 RAM; Ubuntu 18.04
          & Raphael Rochlitz \\
  \emg3d  & \textbf{Laptop}; i7-6600U CPU@2.6 GHz x4; 16 GB of memory,
            Ubuntu 18.04 & Dieter Werthmüller \\
  \petgem & \textbf{Marenostrum4 (supercomputer)}. Intel Xeon Platinum from
            Skylake generation; 2 sockets Intel Xeon Platinum 8160 CPU with 24
            cores each @2.10GHz for a total of 48 cores per node; 386 GB DDR4
            RAM per node; SuSE Linux Enterprise
          & Octavio Castillo-Reyes \\
          \simpeg & \itodo{Todo Lindsey}
          & Lindsey Heagy \\
  %
  \bottomrule
\end{tabularx}}%
%

\imycom[DW]{Raphael: I think here would be the right place? I guess for
  tensor meshes as mine it is simply number of cells (times the anisotropy
  type; for VTI x2). But you know better what it means for \custem/\petgem.
  \imycom[RR]{I think we should introduce the term degrees of freedom (dof),
    equivalent to the size of the SLE, as general measure of our problem sizes.
    If it is somewhat x*Nedges, or related to Ncells or Nnodes, who cares? We
    just need this one number for this paper in my opinion, more details in the
    jupyter notebooks - fine.
  }
}

@@

The customisable electromagnetic modelling Python toolbox custEM was developed for simulating arbitrary complex 3D CSEM geometries with a focus on semi-airborne setups, but it supports also land-based, airborne, coastal, or marine environments. Multiple electric or magnetic field or potential finite-element approaches were implemented as total or secondary field formulations. The finite-element kernel, including higher order basis functions and parallelisation, relies on the FEniCS project \citep{B.SPR.12.Logg, B.SPR.16.Langtangen}. The resulting systems of linear equations are solved with MUMPS \citep{SIAM.01.Amestoy}, which is a very robust but memory consuming choice. Primary field solutions are supplied by the COMET \citep{GEO.20.Skibbe} package.

@@

The 3D CSEM code \emg3d is a multigrid \citep{CMMP.64.Fedorenko} solver for 3D electromagnetic diffusion following \cite{GP.06.Mulder}, with tri-axial electrical anisotropy, isotropic electric permittivity, and isotropic magnetic permeability. The matrix-free solver can be used as main solver or as preconditioner for one of the Krylov subspace methods implemented in SciPy, and the governing equations are discretized on a staggered grid by the finite-integration technique \citep{AEU.77.Weiland}, which is a finite-volume generalization of a Yee grid. The code is written completely in Python using the NumPy and SciPy stacks, where the most time- and memory-consuming parts are sped up through jitted (just-in-time compiled) functions using Numba \citep{LLVM.15.Lam}. The multigrid method is characterized by almost linear scaling both in terms of runtime (CPU) and memory (RAM) usage, and it is therefore a comparably low-memory consumption solver.

The code comes with functionalities to be used directly as a CSEM modeller too, in addition to serve as a solver. As such, it computes the electric field due to an electric source in the frequency-domain, which is its main application (frequency-domain CSEM). However, there are also routines to obtain the magnetic field due to an electric source, to obtain the electric and magnetic fields due to a magnetic source, and to obtain time-domain responses.

Current development is focused on adding basic inversion capabilities, and a further plan is to implement a hook in order to use \emg3d as a frequency-domain CSEM solver within the grander \simpeg framework.

@@

\petgem is a parallel code for frequency-domain 3D CSEM data for marine and land surveys. The high-order edge FE method (HEFEM) is used for the discretization of the governing equations in its diffusive form. This technique provides a suitable mechanism to obtain stable numerical solutions as well as a good trade-off between number of dof and computational effort. The current implementation supports up to sixth-order tetrahedral vector basis functions. Moreover, because the HEFEM belongs to the FE family, the unstructured meshes can be used efficiently on complex geometries, e.g., models with topography and bathymetry.

\petgem permits to locate the source and receivers anywhere in the computational domain (e.g., sediments, seafloor, sea, ground), allowing to analyze the physical environment of the electric responses and how parameters impact them (e.g., frequency, conductivity, dependence on mesh setup, basis order, solver type, among others). Furthermore, \petgem implements a semi-adaptive mesh strategy ($hp$ mesh refinement) based on physical parameters and on polynomial order to satisfy quality criteria chosen by the user. Nonetheless, only Horizontal Electric Dipole (HED) has been implemented.

For the parallel forward modelling computations, a highly scalable MPI domain decomposition allows reducing run-times and the solution of large-scale modelling cases. This strategy is capable of exploiting the parallelism offered by both modest multi-core computers and cutting-edge clusters (e.g., High-Performance Computing architectures).

@@

\simpeg is a modular toolbox for simulations and gradient based inversions in geophysics. Current functionality includes modelling and inversion capabilities for gravity, magnetics, direct current resistivity, induced polarization, electromagnetics (time, frequency, as well as natural sources such as magnetotellurics), and fluid flow (Richards equation). It is a community driven project that aims to support researchers by providing a flexible, extensible framework and to facilitate integration of data sets from a variety of geophysical methods, including joint inversions \citep{GJI.20.Astic}, by providing a common interface to each method. \imycom[DW]{Lindsey: «support researchers» downplays the fact that \simpeg is used by various companies as well\dots}

Meshes and finite volume differential operators are implemented in the \discretize package, which currently includes tensor, OcTree, cylindrical, and logically rectangular meshes. Each mesh type inherits from a common structure and uses the same naming conventions for methods and properties. This allows us to decouple the implementation of a discretized partial differential equation from the details of the mesh geometry and therefore we can write a single implementation of discretized partial differential equation (or set of partial differential equations) in \simpeg that will support all mesh types implemented in \discretize.

The electromagnetics implementations use a staggered grid finite volume approach where physical properties are discretized at cell centers, fields on cell edges and fluxes on cell faces \citep{IEEE.66.Yee}.
%(for an overview of discretizing Maxwell's equations, we recommend \cite{B.SIAM.14.Haber}).
There are two different discretization strategies implemented for Maxwell's equations: (a) the EB-formulation, which discretizes the electric field ($\vec{e}$) on cell edges and the magnetic flux density ($\vec{b}$) on cell faces, and (b) the HJ-formulation, which discretizes the magnetic field ($\vec{h}$) on cell edges and the current density ($\vec{j}$) on cell faces. In both cases the physical properties: electrical conductivity / resistivity and magnetic permeability are discretized at cell centers. Having multiple implementations allows for testing that compares results from each approach, as well as the representation of both electrical and magnetic sources on cylindrically symmetric meshes. \simpeg supports variable magnetic permeability and full-tensor anisotropy for the physical properties.

@@

We computed the responses for two different models to verify that all different 3D codes yield the same electromagnetic responses. The first simple \emph{Block model} consists of a layered, anisotropic background with three embedded blocks. The second model, \emph{Marlim R3D}, is based on a very complex, realistic marine CSEM setup.
